---
title: "Starting with data"
author: "Alexia Cardona & Martin van Rongen"
---


```{r, echo = FALSE, purl = FALSE, message = FALSE}
source("setup.R")
```

<br/>

## Understanding data

```{r, echo=FALSE, purl=TRUE, warning=FALSE}
### Presentation of the survey data
```

To be able to do proper data analyses, it is crucial you understand your data. So before we start doing any form of analyses we will first try to understand the dataset that we will be using throughout this course.  Let us first download the file and have a look at the data.

Thinking back to the structure of our R project, we have a working directory. Within the working directory we can create folders to organise our files.
We are going to download some raw data and it is good practice to keep your raw data separate from other data, because that way you can always refer back to the data that you started with.

Let's create a directory called `data_raw` within our working directory. If you are not sure that you are in your working directory, click on the cog symbol with **More** next to it in the bottom right panel and select **Go to working directory**.

Next, click on **New Folder** and create a folder called `data_raw`. **Note:** Remember that R is case-sensitive, so be consistent. Also, it is best to avoid spaces in names, hence the underscore between `data` and `raw`.


We are now ready to download the data, using the R function `download.file()` to download the CSV file that contains the data.  
```{r, eval=FALSE, purl=TRUE}
download.file(url="https://ndownloader.figshare.com/files/2292169",
              destfile = "data_raw/portal_data_joined.csv")
```
Inside the download.file command, the first entry is a character string with the source URL 
("https://ndownloader.figshare.com/files/2292169"). This source URL downloads a CSV file from 
figshare. The text after the comma ("data_raw/portal_data_joined.csv") is the destination of the 
file on your local machine. 

The dataset has the following columns, with each row holding information for a single animal:

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxon             | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

## Reading in data from a file

Next we need to load the data into R and look at _how_ the data is loaded into R.  We will use `read_csv()` from the `tidyverse` package to load into memory the content of the CSV file. 

You can do this using:

```{r, eval=TRUE,  purl=FALSE}
surveys <- read_csv("data_raw/portal_data_joined.csv")
```

This statement doesn't produce any output because, as you might recall,
assignments don't display anything. It does give you information on how the data was loaded.
Note that some columns are classed as `col_double` (numbers) and others as `col_character` (text).
This can be useful and important information, because it tells you how R views the data. It also helps you check your data. For example, if R would view a column as `col_character` but you know that it should only contain numbers, then you know something has probably gone wrong during the data collection.

If we want to have a closer look at how our data has been loaded, we can visualise the contents of the data frame by typing its name `surveys`:

```{r, eval=TRUE, results=TRUE, purl=FALSE}
surveys
```

The first line of the output shows the **data structure** used to store the data imported into: **`tibble`**. A `tibble` is the main data structure used in `tidyverse`.  You can look at `tibble` as the `data.frame` version of `tidyverse`. The first immediate difference from a `data.frame` is that a `tibble` displays the data type of each column under its name (handy) and it only prints as many columns as fit on one screen (even handier, otherwise it would print 34,786 rows!). 

## Data frames

Data frames are one of the most widely used type of data structure in R.  It is very popular as most of the data is readily available in tabular form and it is the also the data structure used when plotting and performing most analyses in R.

A data frame can be compared to what you would see in an Excel spreadsheet: a rectangular data set.

A data frame is the representation of data in the format of a table where the
columns are vectors that all have the same length. Because columns are
vectors, each column must contain a single type of data (e.g., characters, integers,
logical). For example, here is a figure depicting a data frame comprising a
numeric, a character, and a logical vector.

![](./img/data-frame.svg)

Now that we have loaded our data into R and understand the underlying structure of the data, we can move on to analysing our data!
